---
title: "Paper TITLE"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(readr)
library(stringr)
library(xml2)
library(rvest)
library(tidyr)
library(purrr)
library(maps)
```

Summary/Intro

We chose the dataset from the American Housing Survey, which derives sets of data from 1985 and on. The data set includes categorizing each housing unit by factors such as affordability, income through Fair Market Rent and the Adjusted Median Income. Taking into account variables such as Burden (housing cost divided by its monthly income), certain households with a zero or negative income are input as -1 within the data. The data source was recorded from the American Housing Survey. Certain factors, such as Poverty Income, is based on the Census Bureau's official thresholds. 
 
We chose to analyze the data and ultimately view determines whether housing affordability has increased or decreased over the last few decades. Our initial hypothesis assumes that housing affordability within the United States has decreased due to key, significant events that have occurred within the last few years. The United State’s growing income inequality can be attributed to the current state of housing affordability, and is a clear indicator of the shifting income brackets between who is poor and who is rich. Another major event that could have potentially affected the current state of housing affordability, is the 2007 Housing Market crash that affected the entire country, and even the world. Between all of the foreclosures, loan defaults, and stock market crash, we will closely examine the years between 2007 and 2011 to capture a better sense of the housing market affordability status.




Cleaning data


The "combined_ years" data was read into a dataframe using the "read_ csv" function. Within the read_csv function, the column types were explicitly stated as a string variable (c = character, i = integer, etc.), and the nonsensical integer date (negative numbers) were converted to NAs in the dataset. Then, the combined years dataset was cleaned using the piping methods in dplyr. All the column names were set to lowercase, and all the quotations were removed from character values.
Then, an additional column “ownrent” was added which labeled levels 1 and 2 to own and rent as factors. Then, the data was written to a clean_years csv file. 






```{r, echo = F}
col_types <- "cicciiciiciiidicidicddddddiidddddddddddddddddididididididididididididididicccccccccccccccccccccccccd"
 
clean_years <- read_csv("../data/clean_years.csv", col_types = col_types)

  clean_years %>%
    mutate(per = cut(per, 
                     breaks = c(0, 1, 2, Inf),
                     labels = c("1", "2", "3 or more")))  %>%
    group_by(year, per) %>%  
    # Take the median instead of mean to avoid extreme observations skewing the 
    # result.
    summarise(new_y = median(burden, na.rm = T)) %>%
    ggplot(aes(x = year,   
                      y = new_y,   
                      colour = per)) +  
    geom_line() +
    labs(colour = "Household size",
         x = "Year",
         y = "Median burden")

```

The graph shows the development of median burden (proportion of housing costs relative to household income) from 1985 to 2013. The median burden varies dramatically between households of different sizes. The highest burden through the years has been on households of one person. This is understandable, since people living alone cannot split any costs between other earners.

Surprisingly, households of two people have the lowest burdens overall. Our hypothesis is that this is due to a large proportion of two-earner couples in this group. As housing cost burden is simply a household’s monthly housing cost divided by its monthly income, households where all members are earners have a higher monthly income and a lower resulting burden. In households of 3 or more people, the burden picks up again. This is likely due to a larger amount of children and / or retirees in larger households.

Another interesting thing is the dramatic increase of burden during the buildup to the financial crisis 2000-2007. The most probable exlanation is the housing market price bubble. Once the crisis hit in 2007, the effect of dropping incomes continued to increase burdens until 2011. Only in 2013 data did the burdens start to fall from the unprecedently high levels.

```{r, include = F}
url <- "http://www.mapsofworld.com/usa/usa-maps/united-states-regional-maps.html"
html <- read_html(url)
xpath <- '//table[@class="tableizer-table"]'

regions_tbl <- html %>% 
  html_nodes(xpath = xpath) %>%
  html_table %>%
  .[[1]] 

region_1 <- regions_tbl[2:7,] %>% 
  gather(key = region, value = state, 1:2) %>% 
  slice(1:9) %>%
  mutate(region = 1)
  
region_2 <- regions_tbl[10:15,] %>% 
  gather(key = region, value = state, 1:2) %>% 
  mutate(region = 2)

# Add missing Missouri
region_2[6, 2] <- "Missouri"

region_3 <- regions_tbl[18:27,] %>%  
  gather(key = region, value = state, 1:2) %>% 
  slice(c(-10, -15, -16)) %>%
  mutate(region = 3)

region_4 <- 
  regions_tbl[30:37,] %>%  
  gather(key = region, value = state, 1:2) %>% 
  slice(1:13) %>%
  mutate(region = 4)

regions <- rbind(region_1, region_2, region_3, region_4) %>%
  mutate(state = str_to_title(state),
         region = as.character(region))

median_incomes <- clean_years %>%
  group_by(region) %>%
  summarise(zinc2_median = median(zinc2, na.rm = T))

map_df <- map_data("state") %>%
  mutate(region = str_to_title(region)) %>%
  rename(state = region)  %>%
  # Join with census region information
  left_join(regions) %>%
  left_join(median_incomes)
  

usa_map <- map_df %>% 
  ggplot() +  
  geom_polygon(aes(x=long, 
                   y = lat, 
                   fill = zinc2_median, 
                   group = group)) + 
  coord_fixed(1.3) +
  theme_minimal() +
  scale_x_continuous(name = NULL,
                     labels = NULL,
                     breaks = NULL) +
  scale_y_continuous(name = NULL,
                     labels = NULL,
                     breaks = NULL) +
  labs(fill = "Median income") +
  scale_fill_gradient2(midpoint = 25000)

```

```{r, echo=F}
usa_map
```

To gauge income levels in different areas we used the four census regions in the dataset and joined them with a scraped table from mapsoftheworld.com. Using the table we could connect census regions to state names and thus colour a map of the continental US with census region incomes. We used median incomes instead of mean incomes to avoid skewing the results with outliers.

The household median monthly incomes vary between different census regions from 32200 dollars in the South to nearly 25% higher, 40000, in the West. The Northeast trails closely behind the West with 37267.50 dollars, and the Midwest is closer to the South with a median income of 35000.

While the dataset doesn't have more granular location data, even these four regions are very distinct from each other. The large high-income cities in the Northeast and the West likely contribute to the higher respective median incomes, while more rural Midwest and South are poorer.


```{r, echo = FALSE}
regionNames = c("1" = "Northeast", "2" = "Midwest", "3" = "South", "4" = "West")

```



```{r, echo = FALSE}

cleanyears <- read_csv("../data/clean_years.csv", col_types = "cicciiciiciiidicidicddddddiidddddddddddddddddididididididididididididididicccccccccccccccccccccccccd") %>% 
  select(zinc2, year, region, fmtburden)%>% 
  filter(fmtburden != ".") %>% 
  mutate(fmtburden = str_replace(fmtburden, "Greater than 50%", "50% or More"))

###### Household Income
  ####overall graph
cleanyears %>%  
    group_by_("year", "fmtburden") %>%  
  summarise_(new_y = as.formula(
    paste0("~ median(","zinc2", ",na.rm=T)"))) %>% 
  

  ggplot(aes_string(x = "year", y = "new_y", color = "fmtburden")) + 
  geom_line() + 
  labs(title = "Relationship between Household \nIncome and Burden", 
       y = "Household Income ($)", x = "Year", color = "Burden") 

 

```



```{r, echo = FALSE}


 ### faceted by region 
  cleanyears %>%  
  group_by_("year","region", "fmtburden") %>%  
  summarise_(new_y = as.formula(
    paste0("~ median(","zinc2", ",na.rm=T)"))) %>% 

  ggplot(aes_string(x = "year", y = "new_y", color = "fmtburden")) + 
  geom_line() + 
  facet_wrap(~region, labeller = as_labeller(regionNames)) +
  labs(title = "Relationship between Household \nIncome and Burden", 
       y = "Household Income ($)", x = "Year", color = "Burden") 



```




```{r, echo = FALSE}

cleanyearsmonthly <- read_csv("../data/clean_years.csv", col_types = "cicciiciiciiidicidicddddddiidddddddddddddddddididididididididididididididicccccccccccccccccccccccccd") %>% 
  select(year, zsmhc, region, fmtburden) %>% 
  filter(fmtburden != ".") %>% 
  mutate(fmtburden = str_replace(fmtburden, "Greater than 50%", "50% or More"))

####overall graph

cleanyearsmonthly %>%  

  group_by_("year", "fmtburden") %>%  
  summarise_(new_y = as.formula(
    paste0("~ median(","zsmhc", ",na.rm=T)"))) %>% 
  
  ggplot(aes_string(x = "year", y = "new_y", color = "fmtburden")) + 
  geom_line() + 
  labs(title = "Relationship between Monthly \nHousing Costs and Burden", 
       y = "Monthly Housing Costs ($)", x = "Year", color = "Burden") 


```


```{r, echo= FALSE}

### faceted by region
  cleanyearsmonthly %>%  
  
  group_by_("year","region", "fmtburden") %>%  
  summarise_(new_y = as.formula(
    paste0("~ median(","zsmhc", ",na.rm=T)"))) %>% 

 
  ggplot(aes_string(x = "year", y = "new_y", color = "fmtburden")) + 
  geom_line() + 
  facet_wrap(~region, labeller = as_labeller(regionNames)) + 
  labs(title = "Relationship between Monthly \nHousing Costs and Burden", 
       y = "Monthly Housing Costs ($)", x = "Year", color = "Burden") 


```


```{r, fig.width=12, fig.height=6,echo = F, message=F,error=F, warning=FALSE}

clean_sample <- read_csv("../data/clean_years.csv")
clean_sample %>% filter(fmtstructuretype != ".") %>% 
      # Use NSE
      group_by_("year", "fmtstructuretype") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","increlamipct", ",na.rm=T)"))) %>%  
      # Use NSE
      ggplot(aes_string(x = "year",   
                        y = "new_y",   
                        colour = "fmtstructuretype")) +  
      geom_line() +
  labs(title="Income compared to area's median grouped by house type",
       colour = "Structure Type",
       y="Income percentage of area median income",
       x = "Year")


```

This plot charts a time series showing how the typical income for various types of households such as single family or apartment buildings changed over the years. One major trend is that typical income of all housing types except single family homes compared to their area has fallen, accelerating after 2007. However, incomes of single family homes, barring the housing crash in 2007, have been centered on the area's median income. There are a couple of possible explanations for this disparity. The first is that, with the growth of wages, more people could afford single family homes and thus leave the buildings with higher numbers of units. As wage grew, the area's median income also grew, and so, combined with higher waged workers moving into single family houses, single family homes remained constant while all other types shrunk proportionally. This would also explain the acceleration after the housing crisis, as single family homes became much cheaper temporarily and thus allow more workers to obtain single family housing. The second explanation is that as time went on, supply of housing increased, driving prices down and allowing lower waged laborers to obtain homes. Due to the purchasing power of higher wage laborers however, single family houses, even with their falling prices, were more difficult to obtain as time went on, and thus lower wage workers had to settle for other types of housing. 

```{r, fig.width=12, fig.height=6, echo = F, message=F,error=F, warning=FALSE}

clean_sample %>% filter(assisted == 0) %>% 
      # Use NSE
      group_by_("assisted", "fmtcostmedrelamicat") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","burden", ",na.rm=T)"))) %>%  
      # Use NSE
      ggplot(aes_string(y = "new_y",   
                        x = "fmtcostmedrelamicat",
                        fill = "fmtcostmedrelamicat")) +  
      geom_bar(stat="identity") + labs(title="Burden based on Median housing cost without assistance",
       fill = "Cost as Percentage of Area Median Income",
       y="Burden",
       x = "") + 
  scale_y_continuous(limits=c(0,1)) +
  guides(fill = guide_legend(reverse = TRUE)) +
  coord_flip()
clean_sample %>% filter(assisted == 1) %>% 
      # Use NSE
      group_by_("assisted", "fmtcostmedrelamicat") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","burden", ",na.rm=T)"))) %>%  
      # Use NSE
      ggplot(aes_string(y = "new_y",   
                        x = "fmtcostmedrelamicat",
                        fill = "fmtcostmedrelamicat")) +  
      geom_bar(stat="identity") + labs(title="Burden based on Median housing cost with assistance",
       fill = "Cost as Percentage of Area Median Income",
       y="Burden",
       x = "") + 
  scale_y_continuous(limits=c(0,1)) +
  guides(fill = guide_legend(reverse = TRUE)) +
  coord_flip()


```

The plot shows how burden changes as a result of assistance based on what percentage of the area median income is their monthly housing costs. At initial observation, the burden of those who recieve more assistance is higher than those who recieve no assistance. This clearly makes sense, as those who need assistance are typically people who've lost their job or had some monetary crisis that would reduce their total salary. However, although assistance has around the same effect for most cost brackets, the households that have excessively high costs for their area have vastly higher burdens with assistance than all other brackets. One explanation for this is that those who purchase expensive houses in their area typically have higher salaries, and if they lose those salaries and require assistance they have to support the cost of the house. Also, single family homeowners have significantly less mobility so if they have a financial crisis they cannot easily transition to cheaper housing as opposing to those with rental properties typical of a larger number of units.




```{r}












```




