---
title: "Final Paper"
author: "William Paul"
date: "July 31, 2016"
output: html_document
---

```{r setup, include=FALSE, echo = F, message=F,error=F}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(readr)
library(stringr)
```

```{r, fig.width=12, fig.height=6,echo = F, message=F,error=F, warning=FALSE}
clean_sample <- read_csv("../data/clean_years.csv")
clean_sample %>% filter(fmtstructuretype != ".") %>% 
      # Use NSE
      group_by_("year", "fmtstructuretype") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","increlamipct", ",na.rm=T)"))) %>%  
      # Use NSE
      ggplot(aes_string(x = "year",   
                        y = "new_y",   
                        colour = "fmtstructuretype")) +  
      geom_line() +
  labs(title="Income compared to area's median grouped by house type",
       colour = "Structure Type",
       y="Income percentage of area median income",
       x = "Year")
```
This plot charts a time series showing how the typical income for various types of households such as single family or apartment buildings changed over the years. One major trend is that typical income of all housing types except single family homes compared to their area has fallen, accelerating after 2007. However, incomes of single family homes, barring the housing crash in 2007, have been centered on the area's median income. There are a couple of possible explanations for this disparity. The first is that, with the growth of wages, more people could afford single family homes and thus leave the buildings with higher numbers of units. As wage grew, the area's median income also grew, and so, combined with higher waged workers moving into single family houses, single family homes remained constant while all other types shrunk proportionally. This would also explain the acceleration after the housing crisis, as single family homes became much cheaper temporarily and thus allow more workers to obtain single family housing. The second explanation is that as time went on, supply of housing increased, driving prices down and allowing lower waged laborers to obtain homes. Due to the purchasing power of higher wage laborers however, single family houses, even with their falling prices, were more difficult to obtain as time went on, and thus lower wage workers had to settle for other types of housing. 
```{r, fig.width=12, fig.height=6, echo = F, message=F,error=F, warning=FALSE}
clean_sample %>% filter(assisted == 0) %>% 
      # Use NSE
      group_by_("assisted", "fmtcostmedrelamicat") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","burden", ",na.rm=T)"))) %>%  
      # Use NSE
      ggplot(aes_string(y = "new_y",   
                        x = "fmtcostmedrelamicat",
                        fill = "fmtcostmedrelamicat")) +  
      geom_bar(stat="identity") + labs(title="Burden based on Median housing cost without assistance",
       fill = "Cost as Percentage of Area Median Income",
       y="Burden",
       x = "") + 
  scale_y_continuous(limits=c(0,1)) +
  guides(fill = guide_legend(reverse = TRUE)) +
  coord_flip()
clean_sample %>% filter(assisted == 1) %>% 
      # Use NSE
      group_by_("assisted", "fmtcostmedrelamicat") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","burden", ",na.rm=T)"))) %>%  
      # Use NSE
      ggplot(aes_string(y = "new_y",   
                        x = "fmtcostmedrelamicat",
                        fill = "fmtcostmedrelamicat")) +  
      geom_bar(stat="identity") + labs(title="Burden based on Median housing cost with assistance",
       fill = "Cost as Percentage of Area Median Income",
       y="Burden",
       x = "") + 
  scale_y_continuous(limits=c(0,1)) +
  guides(fill = guide_legend(reverse = TRUE)) +
  coord_flip()
```
The plot shows how burden changes as a result of assistance based on what percentage of the area median income is their monthly housing costs. At initial observation, the burden of those who recieve more assistance is higher than those who recieve no assistance. This clearly makes sense, as those who need assistance are typically people who've lost their job or had some monetary crisis that would reduce their total salary. However, although assistance has around the same effect for most cost brackets, the households that have excessively high costs for their area have vastly higher burdens with assistance than all other brackets. One explanation for this is that those who purchase expensive houses in their area typically have higher salaries, and if they lose those salaries and require assistance they have to support the cost of the house. Also, single family homeowners have significantly less mobility so if they have a financial crisis they cannot easily transition to cheaper housing as opposing to those with rental properties typical of a larger number of units.