---
title: "APL_region"
author: "Sriharsha Kondapally"
date: "July 30, 2016"
output: html_document
---


```{r}
library(ggplot2)
library(readr)
library(stringr)
library(dplyr)

col_types = "cicciiciiciiidicidicddddddiidddddddddddddddddididididididididididididididicccccccccccccccccccccccccd"

clean_years <- read_csv("../data/clean_years.csv", col_types = col)

clean_years %>%
  group_by(year, region) %>%  
  # Take the median instead of mean to avoid extreme observations skewing the 
  # result.
  summarise(new_y = median(aplmed, na.rm = T)) %>%
  ggplot(aes(x = year,   
                  y = new_y,   
                  colour = region)) + geom_line() +
  labs(x = "year", y = "Median Adjusted Income (per persons)", 
       title = "APLmed by region")
```

This plot compares APLmed by year for the four regions in the United States (Northeast, South, Midwest, West). It is clear that the Northeast and the West are the most expensive regions. This is makes sense given that the northeast and West are by far the most productive (by GDP) regions in the United States. 

The median adjusted income increases steadily (roughly linearly) over time. There is a decrease in income around 2010, correlated to the economic recession with a 2 year lag between the Northeast and the rest of the regions.The reason for this lag may be due to the fact that the financial center of the United States lies in the northeast, around New York City. So, when the 2008 crisis occurred the repercussions were immediately felt in the Northeast (Wall Street and Main Street- which refers to the Washington D.C. economy) while the other regions took some time to correct for the downturn in the financial market. 

```{r}
library(ggplot2)
library(readr)
library(stringr)
library(dplyr)

col_types = "cicciiciiciiidicidicddddddiidddddddddddddddddididididididididididididididicccccccccccccccccccccccccd"

clean_years <- read_csv("../data/clean_years.csv", col_types = col)

clean_years %>%
  group_by(year, region) %>%  
  # Take the median instead of mean to avoid extreme observations skewing the 
  # result.
  summarise(new_y = median(ablmed, na.rm = T)) %>%
  ggplot(aes(x = year,   
                  y = new_y,   
                  colour = region)) + geom_line() +
  labs(x = "year", y = "Median Adjusted Income (per persons)", 
       title = "ABLmed by region")
```

The trends of the graphs here are similar to the ones in APLmed, however ABLmed signifies the median adjusted income for the number of bedrooms in a household's living quarters. 

It is interesting to note that the disparity between the incomes increases between the Northeast and WEstern regions and the Midwest and South. This is liekly because those living in the Midwest and South ten dto live in larger homes, housing more bedrooms on average, relative to their incomes due to the higher housing affordability. However, this should mean that their incomes go 'further,' as they have more purchasing parity given the lower cost of living compared to the Northeast and the West. This is not shown in the graphs, which indicate that the income levels are significantly different. However, if housing affordability was factored into the calcualtion, the income disparity would be far less if not almost nonexistent. 

The lag in the downward trend after 2008 still occurs in the plots.
